Conservator's Corner

### 2020-02-24 pnsac/newsletter/article_MemberInterview.tex
Draft updates for Feb-Mar 2020.

\title{Our Members}
\author{Interview with Austin J Timmins}

@@

My aviation experience includes 36 years service in the RCAF and nine
years with the Canadian Aeronautics and Space Institute. In the RCAF I
completed radio navigator training, which included communications,
navigation and air gunner. I later qualified as an airborne
interception navigator. My operational experience includes air
transport, search and rescue and training. My assignments included
Navigation Leader, Aircrew Leader, Squadron Commander, Air Command
Navigator and Deputy Commander (COO) of the Air Transport Group. I
logged 9500 hours, mostly on the following aircraft: Beechcraft,
Dakota, North Star, Mitchell, Yukon, Hercules, and Boeing 707. My
staff assignments in Europe included movement control (traffic
management) for 1 Air Division and a tour at The NATO Defense College
in Rome.

On retirement from the RCAF I participated in a Transport
Canada A Base review of several airports. I then secured full
employment with Canadian Aeronautics and Space Institute as Executive
Director, retiring in 1995.

@@

My involvement with efforts to conserve North Star 17515 predates
Project North Star. Ottawa area members of the 426 Thunderbird Squadron
Association, mostly retired North Star aircrew, met with the National
Aviation Museum staff to express their concern about the state of the
North Star and proposed volunteer involvement in a preservation effort.
These expressions of concern and interest led to a meeting between the
Association representatives, where General Adamson presented a proposal
for the restoration of the aircraft. It was rejected but assurance was
given that the North Star would be first priority when the new hangar
was completed Also a North Star Trust Fund would be established. Robert
Holmgren, an observer at the meeting, then proceeded to develop a
proposal for volunteers to restore the North Star.

@@

I have been involved from the beginning of efforts to establish PNS,
first as a member of a Steering Committee, led by Robert Holmgren,
later, Vice President and then President of PNSAC. My contacts in
aviation provided much needed advice and support, including access to
initial funding for PNSAC.

@@

The highlight of my involvement has to be the establishment of PNS and
PNSAC.

@@

PNS is now in its 16th year. It has been a work in progress with many
bumps along the way. I spent a lot of time and effort, along with other
members of the Executive, evening out these bumps. Establishing and
maintaining effective channels of communication with CASM and CSTM
required constant attention. Much effort was directed to fund raising,
publicity, and, finding volunteers. In the beginning, every date
brought a new challenge.

>>>>

\title{Our Members}
\author{Interview with Garry Dupont}

@@

I graduated from the Confederation College Aircraft Maintenance course
in the Spring of 1975. After returning to my home town in the Red Lake
District I gained employment with Red Lake Seaplane Service. After
completing the course I had to apprentice for 18 months before writing
the final exam for my Aircraft Maintenance Engineers' License. I
successfully wrote my final exam in December 1976 and obtained my AME
certification. In February 1977 I was hired by Perimeter Aviation in
Winnipeg Manitoba. There I obtained my Turbine Engine and
Pressurization Aircraft endorsements.  I started my 30 year career with
the RCMP Air Services Branch in February 1981. My career took me to
Edmonton Alberta, Goose Bay Labrador, Winnipeg Manitoba then finally to
Ottawa. I finished my career as Quality Assurance Manager.

@@

I attended the very first meeting held for Project North Star. This was
at my wife's suggestion that it would be a good retirement pastime.

@@

While I was employed I worked an evening shift every 4 weeks so I would
come in to CASM in the morning for a few hours. Upon my retirement Mike
Irving had me work in the engine shop, which I have been doing since my
retirement. I also have been on the board of directors for a number of
years.

@@

The biggest highlight for me was installing number 2 engine. It was
about half completed when I started in the engine shop. This is the
largest engine I have ever installed in my entire career.

@@

One of the most challenging parts was learning how the engine worked
with all its systems. I found a RCAF training manual in the office and
took it home. Over several months of reading I started to understand
what made this monster of old technology tick. Another challenge was
learning the British Hardware system and tools. A further challenge was
reading  the old manuals which were written before they were
standardized by the ATA chapter system which all aircraft manuals
follow today. Learning the art of conservation was also a  challenge
but was made easier by the guidance of Mike Irving.

### 2020-02-24 pnsac/newsletter/article_PresReport.tex
Draft updates for Feb-Mar 2020.

These are exciting times to be a member of the Project North Star Association.
We  have recently taken part in the 2018 Doors Open Ottawa. This is a major
event for the Museum and PNSAC volunteers were able to make a significant
contribution to the success of the two open days. Doors Open has become second
only to Canada Day in its importance to the Museum. (more on Open Doors 2018
later in this issue).

A few days before Canada Day the aircraft was brought outside. Unfortunately,
the day itself was extremely hot and the number of visitors was well down from
previous years. Nonetheless our volunteers did a yeoman service in talking to
the public about our restoration work.

Although showing the aircraft to the public is very important to us, our main
purpose is to restore the North Star to museum condition. Much of our work is
detailed, painstaking and unspectacular such as restoring a small part of the
aircraft which can take many days or weeks. At other times making progress is
very visible. We are in such a situation now.

At the end of May we had one of our regular meetings with Chris Kitzan (Director
General of CASM) together with Neil Raynor (VP of PNSAC), Cedric St-Amour (CASM
Volunteer Coordinator) and Reg Demers (Our CASM supervisor There was a full
exchange of ideas, suggestions for bringing  the public into the Museum and ways
in which PNSAC could be involved in these initiatives and events. The meeting
ended with me having a full sheet of notes covering future planning, major
restoration work and how we can make sure we recruit younger people both as
members of PNSAC and where possible active Restoration Volunteers.

A meeting like the one above only happens when an organization is thriving, and
everybody is excited about the future. During the last 15 years PNSAC has
demonstrated that it is capable delivering very high-quality restoration work.
We have already accumulated over 80,000 hours of work and during this time we
have become a trusted partner of the Museum. Our website is now operational
again and in the next few weeks we expect to be able to update it and make it
more relevant to where we are with the project now. One of the major initiatives
taking place is to develop a diagrammatic history of the work done on the North
Star restoration and a timeline for future work on the aircraft.

In closing, I once again urge Association members to become active volunteers.
There are many volunteering opportunities both for those with major aviation
experience to those who can only volunteer with us on occasions like Canada Day
or can provide help for our many behind-the-scenes activities.

Our next Members' Meeting is scheduled for
Saturday, September 29, 2018. Do reserve this date in your calendar if you can
be in the Ottawa area on that day. The meeting will be held at the Canada
Aviation and Space Museum. More details will follow later.

>>>>

Since the last issue of the NStar Chronicle we have taken part in the
70th Anniversary celebrations of the first non-stop transcontinental
flight across Canada which was flown by a North Star. We partnered with
the Museum for this event and were very pleased to welcome
representatives from the RCAF who joined us for the day. After the
formal part of the celebration, when impressive plaques of the aircraft
made by our member Chris McGuffin, had been presented to the RCAF and
the Museum Director General, Chris Kitzan, everybody moved to the
Reserve Hangar and the aircraft was opened to our association members,
guests and in the afternoon to 90 members of the public who had
pre-booked a tour. R\'{e}jean(R\'{e}j)Demers, the Museum Special Project
Manager, guided the tours assisted by several of our members. Towards
the end of the tour session someone in the main Museum activated the
fire alarm. The Museum was promptly evacuated, and the public was
recommended to go to the Reserve  Hangar until the fire issue had been
dealt with (it turned out to be a false alarm). Quite unexpectedly R\'{e}j
and the guides found themselves surrounded by many Museum visitors who
had not signed up for the North Star tour. Unperturbed by this event,
Réj and the guides welcomed the unscheduled visitors and gave them an
impromptu tour of the aircraft. Both our treasurer and our membership
secretary were pleased with the days' result!

As I write this the days are getting longer and R\'{e}j is planning work to
be done on the aircraft when it is moved outside onto the ramp in the
spring. Each year we do work on the aircraft which cannot be done
inside the Reserve Hangar during the winter. We also prepare the
aircraft for display when, once again, we open the aircraft to enable
the public to view progress made in the last year.

This year we are planning two major display events. The first will be
on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
for our members to meet the public, many of whom have an interest in
aviation. Our second main event, on Canada Day, is enjoyable for all
the members who sign up to be part of the team meeting the public.
There are always large crowds, many of them family groups, If any of
you reading this article would be interested in joining our members for
either of these events please contact me (my co-ordinates are on the
back page of this issue of the Chronicle)

Shortly before Christmas a major milestone, in the  restoration of the
North Star, was reached. Work on Merlin engine \#4 was completed and it
was dragged on its cradle by the Museum forklift truck to the
Restoration hangar for later mounting on the aircraft to join the other
three restored engines. Chris McGuffin put some video clips onto our
Facebook page of the engine being dragged through the Museum parking
lot and into the hangar. Watch for more of our Facebook messages since
Chris frequently posts new items.

In the late fall we welcomed the appointment of Erin Secord as the
Museum's new Manager Conservation Services. She is responsible also for
conservation of the collections at the Science and Tech Museum. Erin
has now been able to join us at two of our Members' meetings.

Finally, we are now working to update the Memorandum of Understanding
between PNSAC and the Museum.  Much has changed since the original
agreement was signed in May 2006.

Our next Members' Meeting is scheduled for Saturday, June 20, 2020
which will also be our AGM. Do reserve this date in your calendar if
you can be in the Ottawa area on that day. The meeting will be held at
the Canada Aviation and Space Museum. More details will follow later.

### 2020-02-24 pnsac/newsletter/article_ProgReport.tex
Draft updates for Feb-Mar 2020.

\title{Conservator's Corner}
%\subtitle{PNS 2016 Status Update}
\author{R\'{e}jean Demers}
\author{Restoration Project Manager}

@@

\section{Zen and the Art of Aircraft Restoration.}

Through my daily experience over the past two years, working with volunteers, I
have asked myself the following question: What makes a volunteer dedicated to a
project? The answer is not a straightforward one. This article explores the
subject, much like Pirsig did in his book about metaphysical motorcycles; Zen
and the Art of Motorcycle Maintenance.

We strive for quality. As restoration work continues on the North Star, we get
absorbed into the weight and breadth of the subject matter. This is the kind of
project that demands the level of commitment found in monasteries, dojos and
Paratrooper units. Commitment comes from dedication and reward. Quality is
obtained through application of commitment, producing results which may be
enjoyed. The following effort in writing has been made to define such elements,
those necessary to sustain the project.

A voracious appetite for detail. Precise facts and data build foundations.
Foundations of information that erode over time. Lost to floods, fires, or
worst; thieves, vandals and neglect. Information is our firmament that keeps us
grounded in our work. Daily worksheets are filled with steps taken to satisfy a
condition. Drawings are made to translate reality to paper. These words,
drawings and tables form hard to translate time lines. Coloured bars overtaking
each other, running in thousand hour blocks of volunteer contributions. These
life hours, cannot be overlooked, as a restoration project advances through the
years.

Inorganic compounds risk distraction. Focus is important when avoiding the
ethereal draw of solvents. A volunteer scrubber, washing away years of residue,
layered upon a finish that once gleamed with pride. Slowing the return to dust,
so feared and fought by the living. A process to follow and adhere to. The
methodical dismantling and reconstruction of an object. Return to form, that
which was almost indiscernible. The result of all his work is not merely a
shiny piece of aluminium panelling; He has travelled through time.

More than just an object, we look to a hidden past. Beyond these material
trappings, lay the history of Canadair North Star 17515 and her sisters.
Examples of service are easily found in books and photos. What proves a
challenge is explaining how a bumblebee becomes a preserved specimen in an
engine, how graffiti made during heavy maintenance is the only evidence of a
man's past. Where the dirt under floorboards, came from countries that no
longer bear the name she visited.

The static, prevents complete appreciation. She's more than 45,000 lbs. of
metal. Superficial understanding can only consider her 39 years an outdoor
relic. Likewise, so would a skin deep treatment of a last remaining artefact of
this type. What impact she had before the stillness of the seasons is not lost
to those who look further.

A biological experience in personal evolution. The volunteer has developed new
muscles and sinew adapted to producing polished surfaces. An eye for
microtexture and grain direction has developed. Sympathetic scars are
accumulated through careful little sacrifices. Navigating the disruptions of
capricious human needs such as rest, nourishment and leisure. Through the
application of labour, a piece of history is preserved.

Disappearing horizons. Meditative expanses of aluminum, exhibiting varieties of
corrosion rarely encountered in a lifetime's career. Cavernous interior spaces,
where cargo of all sorts was bound for ports beyond. Flight control surfaces
that have pressed through all shades of dusk, emerging in the dawn of a new
day. Now removed from the ship, these weathered, beaten, bruised remnants of a
glorious past await restoration. A new horizon forms, some many years away,
known to those who are on-board for the long run.

Healthy obsessions. Volunteer restoration technicians take home an experience.
They take to heart the results of their labour, in forming a product that is
also intangible. The mind has strengthened, forming resolves that assist,
understand and process the elements of restoration work. As historians,
fanatics and unwitting library wanderers soon discover; The story of air
transport, in its beginning, is far from just a few obscure paragraphs in
Canada's ongoing history. It lives through those people who continue writing
it, preserving the physical, extending beyond.

>>>>

\title{Conservator's Corner}
%\subtitle{PNS 2016 Status Update}
\author{Bruce Gemmill, R\'{e}jean Demers, and Garry Dupont}
%\author{Restoration Project Manager}

@@

%\section{Zen and the Art of Aircraft Restoration.}

The past year was a very busy one for the North Star restoration crew. 
In the Engine Shop, under the careful watch of Crew Chief Garry Dupont,
members Charles Baril and Richard Lodge worked to fit the final pieces
on engine 4 including; all the cowl frames and panels, main and
auxiliary radiator flaps under the engine, the engine air intake duct
and associated actuators. All the hard work in setting up sheet metal
repairs to the cowl panels paid off.  However, fitting the large steel
exhaust panels around the 12 exhaust stacks was particularly
challenging. It's no longer a secret that ratchet straps and wooden
2"x4" blocks have become part of the standard North Star tool kit.

While that work was underway, Jacques Roy, Peter Scartozzi and Charles
Baril were busy stripping equipment, pipes, hose and cables from inside
the nacelle where engine 4 will soon be installed.  Will Assad and Neil
Raynor also helped with this. After equipment removal, many hours were
spent carefully cleaning and removing oil and corrosion prior to
painting.  In some areas a small bead blasting gun was used to remove
rust from nuts and bolt heads that could not be taken out for cleaning.

Most notably were efforts by Charles Baril, who spent many months
inside the nacelle; cleaning, inspecting and masking off for painting.
The work took so long that special heating ducts and insulation needed
to be positioned around the nacelle to keep it warm enough for Réjean
Demers and Bruce Gemmill to apply a coat of primer and silver paint.
(insert photo of North Star with orange tarp). Soon after the paint was
applied, the North Star was moved inside for the winter. This occurred
November 8th, just 3 days before our first snow storm of the winter.

While all that work was underway, the Airframe Crew was busy on the
structures work.  Peter Trobridge, Michel Cote and Robert Desjardins
fabricated and installed several replacement parts for main cabin
airframe assemblies that had suffered from years of corrosion.  The
crew accomplished a complete review of necessary repairs from front to
back of the main cabin. A hole was patched near the door to the crew
area, at forward bulkhead (station) 301. This offered Chris McGuffin an
intro to sheet metal repairs. Similar corrosion damage around the
windows required parts shaped on wooden jigs made in the shop. Through
careful planning and creative solutions, repairs were completed at STA:
360, and 381. New repairs carried out at starboard STA: 460, 480, 501,
541, 560, 621 and currently, floor repairs in the loading area near the
cargo doors.

The Carpentry Crew, led by Chris McGuffin measured, cut and trimmed new
plywood panels for the interior walls of the main cabin.  These were
painted green to match the original colour scheme, then set aside until
for later installation.  Next came the floor panels.  Richard Houle and
John Makadi assessed these, according to the amount of damage caused by
wear and tear, as well as leaks around windows and doors.  Some panels
were salvaged with a few minor repairs, then painted. These are now
ready for installation.  Several others were so badly damaged that
completely new panels needed to be made. (Photo of wall and floor
panel)

Finally, work started on the third group of panels.  These floor panels
had significant damage to one or two edges.  The damaged areas need to
be cut away and a new piece added by bevelling the adjoining pieces
using what is called a "scarf joint" so the adhesive bond is strong.  A
special jig was constructed to do this work, allowing an electric
router to run at an angle to produce the joint. A Resorcinol
Formaldehyde adhesive will assure a long-lasting repair.

Many of the volunteers got to work on the large collection of
accessories that had been removed from nacelle \#4.  This included
pumps, plumbing, fluid and air lines, control cables, bell cranks and
pulleys, electrical harnesses, valves and filter canisters. Each needed
assessment, cleaning, repair, polishing or painting as well as
documentation and photographs of the work undertaken. (photo of fuel
filter, junction box).

Phil Chrysler and Brian Cole went to work on the many fairings and
access panels that also required cleaning and polishing to be ready to
be placed back on the aircraft. A special mention here for Ted Devey,
who, before retiring from the shop had his hand on a few cowl panels
and nacelle components.

Several special projects were also undertaken during the year.  John
Makadi removed the nose radome to determine what radar equipment had
been removed, in case we are able to find a source for missing
equipment.  We removed the upper rotating navigation beacon to fix a
persistent leak, this necessitated a twin tower scaffold assembly to
safely reach the top of the fuselage. Neil Raynor and Will Assad
overhauled a propeller stand the Museum had not used for many years. 
It got a fresh coat of paint and new wheels and tires, and is now
holding two North Star propellers, until they can be installed on the
appropriate engine. (photo of prop stand, two propellers). Michael Hope
assisted in removal of Prop. \#1 and assembly of soon-to-be-installed
Prop. \#4. Phil Chrysler and Brian Cole fabricated cables and special
fittings to secure the landing gear so the aircraft could be safely
displayed. Other duties  such as; an inventory of replaced / restored
and spare components were carried out.  All of these were packed and
stored in a new location of the Reserve Hangar.

Number \#4 engine has now  been moved to the Reserve Hangar. The crew
are prepping the oil tank for clear coat then it will be assembled and
ready to install. Once this is completed  the plan is to carry out some
deep cleaning and renovations in the shop starting  by  relocating the
varsol cleaning tank. The paint shaker has been repaired and the base
has been painted ready for assembly. The paint shaker will be relocated
where the varsol tank was. The old roll away tool box has been replaced
by a new Snap-On unit. There have been lots of comments on the orange
colour  but the price was right.

Currently, work is on-going to complete the rear (main) cargo door and
quilted liner installation. Nacelle \#4 installation of restored
components is advancing in small steps, due to the complexity of
interrelated systems and locations. Scaffolding is in place, over the
forward fuselage area above the cockpit to assess air-frame
requirements and treatment methods to be used on exterior hull
surfaces. Restoration records are also being reviewed by Conservation
Special Project Manager Réjean Demers. The Work Order / Task Card
format is lending well to Museum information uploading tools. Efforts
are underway to recruit two to four new shop volunteers, and R\'{e}j
invites any new applicants referred by PNSAC to apply.
